---
output: github_document
bibliography: inst/ref.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# causalXtreme

<!-- badges: start -->
[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Travis build
status](https://travis-ci.com/nicolagnecco/causalXtreme.svg?branch=master)](https://travis-ci.com/github/nicolagnecco/causalXtreme)
[![codecov](https://codecov.io/gh/nicolagnecco/causalXtreme/branch/master/graph/badge.svg)](https://codecov.io/gh/nicolagnecco/causalXtreme)
[![R-CMD-check](https://github.com/nicolagnecco/causalXtreme/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/nicolagnecco/causalXtreme/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of causalXtreme is to provide an interface to perform causal discovery in linear structural equation models (SEM) with heavy-tailed noise. For more details see @gne2019 [https://arxiv.org/abs/1908.05097].

## Installation

<!-- You can install the released version of causalXtreme from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("causalXtreme") -->
<!-- ``` -->

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("nicolagnecco/causalXtreme")
```
## Example

Let us generate 500 observations from a SEM of two Student-t variables, $X_1$ and $X_2$, with 1.5 degrees of freedom (i.e., heavy-tailed).
When the function `simulate_data` is called with the default values, it returns a list containing:

  * the simulated `dataset` represented as a matrix of size $n \times p$. Here, $n = 500$ is the number of observations and $p = 2$ is the number of variables,
  * the underlying directed acyclic graph (DAG) represented as an adjacency matrix `dag` of size $p\times p$.

```{r sem}
library(causalXtreme)

# basic example code
set.seed(1)
sem <- simulate_data(n = 500, p = 2, prob_connect = 0.5,
                     distr = "student_t", tail_index = 1.5)
```

At this point, we can look at the randomly simulated DAG.

```{r dag}
sem$dag
```

We interpret the adjacency matrix as follows. Loosely speaking, we say that variable $X_i$ causes variable $X_j$ if the entry $(i, j)$ of the adjacency matrix is equal to 1.
We see that the first variable $X_1$ causes the second variable $X_2$, since the entry $(1, 2)$ of the matrix `sem$dag` is equal to 1. We can plot the simulated dataset.

```{r plotdata, echo=TRUE}
plot(sem$dataset, pch = 20,
     xlab = "X1", ylab = "X2")
```

At this point, we can estimate the causal direction between $X_1$ and $X_2$ by computing the *causal tail coefficients* $\Gamma_{12}$ and $\Gamma_{21}$ [see @gne2019,  Definition 1].

```{r causaltailcoeff}
X1 <- sem$dataset[, 1]
X2 <- sem$dataset[, 2]

# gamma_12
causal_tail_coeff(X1, X2)

# gamma_21
causal_tail_coeff(X2, X1)

```

We see that the coefficient $\Gamma_{12} \approx 1$ (entry $(1, 2)$ of the matrix) and $\Gamma_{21} < 1$ (entry $(2, 1)$ of the matrix). This is evidence for a causal relationship from $X_1$ to $X_2$.

We can also run the *extremal ancestral search* (EASE) algorithm, based on the causal tail coefficients [see @gne2019, Section 3.1]. The algorithm estimates from the data a *causal order* of the DAG.

```{r ease}
ease(dat = sem$dataset)
```

In this case, we can see that the estimated causal order is correct, since $X_1$ (the cause) is placed before $X_2$ (the effect).

## References
