---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# causalXtreme

<!-- badges: start -->
[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Travis build
status](https://api.travis-ci.org/nicolagnecco/causalXtreme.svg?branch=master)](https://travis-ci.org/nicolagnecco/causalXtreme)
<!-- badges: end -->

The goal of causalXtreme is to provide an interface to perform causal discovery in linear structural equation models (SEM) with heavy-tailed noise. For more details see the paper "Causal discovery in heavy-tailed models" from Gnecco, N., Meinshausen, N., Peters, J., and, Engelke, S.

## Installation

<!-- You can install the released version of causalXtreme from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("causalXtreme") -->
<!-- ``` -->

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("nicolagnecco/causalXtreme")
```
## Example

Let us first generate a SEM with two Student-t variables with 1.5 degrees of freedom (i.e., heavy-tailed).

```{r sem}
library(causalXtreme)
## basic example code
set.seed(1)
sem <- simulate_data(n = 500, p = 2, prob_connect = 0.5,
                     distr = "student_t", tail_index = 1.5)
```

Let us investigate the randomly generated directed acyclic graph (DAG) induced by the SEM.

```{r dag}
sem$dag
```

We see that the first variable causes the second variable, since the entry $(1, 2)$ of the matrix `sem$dag` is equal to 1. We can plot the simulated dataset.

```{r plotdata, echo=FALSE}
plot(sem$dataset)
```

At this point, we can compute the *causal tail coefficients* between the two variables $X_1$ and $X_2$.

```{r causaltail}
causal_tail_matrix(dat = sem$dataset)
```

We see that the coefficient $\Gamma_{12} \approx 1$ (entry $(1, 2)$ of the matrix) and $\Gamma_{21} < 1$ (entry $(2, 1)$ of the matrix). This is evidence for a causal relationship from $X_1$ to $X_2$.

We can also run the *extremal ancestral search* (EASE) algorithm, based on the causal tail coefficients. The algorithm estimates from the data a causal order for the DAG.

```{r ease}
ease(dat = sem$dataset)
```

In this case, we see that the estimated causal order is correct, since the cause $X_1$ is placed before its effect $X_2$. 
